Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-05 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: small edits on qeft
hs: edits in results section

Using this parameterized waveform model, which we will call $\pSEOB$ in this
paper, we infer bounds on our beyond-GR parameters for specific cases of
modified gravity theories presented Sec.~\ref{sec:review_theories}.
%
We detail these results in Sec.~\ref{sec:results}.
%
% \hs{Alessandra, Abhirup, should we call this waveform differently to distinguish it from your previous papers?}
%
%In this paper, we will restrict ourselves to the $k=(2,2,0)$ QNM and corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$. Assuming a prior which is uniform on the beyond-GR parameters, $(\ell, \{\delta \omega_k^{(j)}\},\{\delta \tau_k^{(j)}\})$, we employ a Bayesian framework to stochastically sample over the parameter space using the \texttt{LALInference} algorithm.~\cite{lallsuite}.
%
%The bounds on these beyond-GR parameters, for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}, are detailed in Sec.~\ref{sec:results}.

@@

Our result contrasts with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm GB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB provides an example of a theory in which the inspiral portion
of the signal is more constraining then the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_f$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.~\hs{This is a general `limitation' of ringdown test of higher-curvature theories in present detectors.
Point this out in the conclusions after all specific theories are discussed.}
%
As we will next, the situation is the opposite for dCS gravity.

@@

In Fig.~\ref{fig:cEFT_cdf} we show the CDF for
both events, calculated as already described for the EdGB and dCS cases.
%
We see that curves are very similar to those of dCS gravity for GW200129 (cf.
bottom panel in Fig.~\ref{fig:cEFT_cdf}).
%
Moreover, we find that for both events both the EFT~\eqref{eq:eft_bound} and
ParSpec~\eqref{eq:parspec_bound} bounds are satisfied for $\mathfrak{m} = M_f$
and $N_{\rm max} = 1/2$, allowing us to obtain the combined bound
%
\begin{equation}
    \ell_{\rm cEFT} \leqslant 51.3~\textrm{km},
\end{equation}
%
at 90\% credibility. As also happened for our study for dCS, the find that for
the cubic EFT that the ParSpec bound is violated by the $n = 1$ corrections to
the QNMs, implying that we cannot use the associated posterior distributions on
$\ell_{\rm cEFT}$ to draw any meaningful constraint on this parameter.
%
We summarize our results in Table~\ref{tab:summary_ceft}.

@@

At last, let us consider the quartic EFT of GR.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR
%
We find that both the ``EFT''~\eqref{eq:eft_bound} and ``ParSpec''~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_f$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km}\,.
\end{equation}
%

In this theory, we considered only $N_{\rm max} = 0$. We found that the
inclusion of the spin corrections yields waveforms which for $\ell_{\rm qEFT} \gtrsim 50$~km
with ringdowns portions of the larger than the inspiral-plunge part \dots
%
To avoid this issue we lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but at doing so we obtained which
were essentially uniform. Thus, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_cdf_varying_threshold.pdf}
\caption{
}
\label{fig:qEFT_cdf}
\end{figure}

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
}
\label{tab:summary_qeft}
\end{table}

>>>>

Using this parameterized waveform model, which we will call $\pSEOB$ in this
paper, we infer bounds on our non-GR parameter $\ell$ for specific cases of
modified gravity theories presented Sec.~\ref{sec:review_theories}.
%
We detail these results in Sec.~\ref{sec:results}.
%
% \hs{Alessandra, Abhirup, should we call this waveform differently to distinguish it from your previous papers?}
%
%In this paper, we will restrict ourselves to the $k=(2,2,0)$ QNM and corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$. Assuming a prior which is uniform on the beyond-GR parameters, $(\ell, \{\delta \omega_k^{(j)}\},\{\delta \tau_k^{(j)}\})$, we employ a Bayesian framework to stochastically sample over the parameter space using the \texttt{LALInference} algorithm.~\cite{lallsuite}.
%
%The bounds on these beyond-GR parameters, for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}, are detailed in Sec.~\ref{sec:results}.

@@

We can compare this conclusion with that of Ref.~\cite{Carullo:2021dui}, which obtained $\ell_{\rm GB} \lesssim 35$~km,
by setting $p=4$, \emph{but not} including theory-specific QNM information on $\delta \omega^{(n)}$ and $\delta \tau^{(n)}$.
%
Our results provide a concrete example of the importance of including
theory-specific QNM calculations information into the parameter estimation
and how this can dramatically change the outcome of the results.

Let us also contrasts our results with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm GB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB provides an example of a theory in which the inspiral portion
of the signal is more constraining then the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_f$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.~\hs{This is a general `limitation' of ringdown test of higher-curvature theories in present detectors.
Point this out in the conclusions after all specific theories are discussed.}
%
As we will next, the situation is the opposite for dCS gravity.

@@

In Fig.~\ref{fig:cEFT_cdf} we show the CDF for both events, calculated in the
same way as already described for the EdGB and dCS theories.
%
We see that curves are very similar to those of dCS gravity for GW200129 (cf.
bottom panel in Fig.~\ref{fig:cEFT_cdf}).
%
Moreover, we find that the EFT~\eqref{eq:eft_bound} and
ParSpec~\eqref{eq:parspec_bound} bounds are satisfied for both events both when
$\mathfrak{m} = M_f$ and $N_{\rm max} = 1/2$.
%
This allows us to place the combined bound of
%
\begin{equation}
    \ell_{\rm cEFT} \leqslant 38.2~\textrm{km}\,,
\end{equation}
%
at 90\% credibility. As also happened for our study for dCS, the find that for
the cubic EFT that the ParSpec bound is violated by the $n = 1$ corrections to
the QNMs, meaning that we cannot use this case to draw any meaningful
constraint on this parameter.
%
We summarize our results in Table~\ref{tab:summary_ceft}.

@@

Let us now consider the quartic EFT of GR as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR
%
We find that both the ``EFT''~\eqref{eq:eft_bound} and ``ParSpec''~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_f$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km}\,.
\end{equation}

In this theory, we considered only $N_{\rm max} = 0$. We found that the
inclusion of the spin corrections yields waveforms which for $\ell_{\rm qEFT} \gtrsim 50$~km
with ringdowns portions of the larger than the inspiral-plunge part \dots
%
To avoid this issue we lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we obtained posteriors
which were essentially uniform. Thus, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_cdf_varying_threshold.pdf}
\caption{Similar to Fig.~\ref{fig:edgb_cdf}, but for the quartic EFT of GR.
We find that similarly to what happens in dCS gravity and the cubic EFT of GR, we
can place a constrain on $\ell_{\rm qEFT}$ when $\gm = M_f$ for both events at 90\% credibility.
}
\label{fig:qEFT_cdf}
\end{figure}

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond to 90\% credible values.
}
\label{tab:summary_qeft}
\end{table}
